Counting And Confusion: Bayesian Rate Estimation With Multiple Populations

### 2013-02-19 many.tex
Implemented Tom Dent's comments from LVC review.

The task of estimating rates of events when a mixture of foreground
and background events is present in data is a common one in physical
and astrophysical applications.  This problem comes up, among others,
in gravitational-wave data analysis
\cite[e.g.,][]{s5-highmass,s6-lowmass} and in astronomical
observations of a field of objects of mixed provenance
\cite{DeGennaro2009}.  In this paper, we introduce a robust formalism
for estimating event rates from the data when the shape of foreground
and background distributions are known (or parameterized), but the
provenance of individual events as either background or foreground is
unknown.

@@

Two alternative approaches to rate estimation have been suggested and
are commonly used.  One, known as the \emph{loudest-event statistic}
\cite{Biswas2009}, uses only the information from the highest-ranked
event in the data to infer the rate distribution.  This approach has
been used successfully \cite{s5-highmass,s6-lowmass} when the number
of loud foreground events is small (typically zero or one) to obtain
upper limits on foreground rates.  However, the loudest-event
statistic ignores all events except the loudest one, and so suffers
from an unnecessary loss of information; therefore, we expect it to
yield a much larger variance than strictly necessary when multiple
events are present in the data.

@@

@@

For the usual $k=1$ case \citep{Biswas2009}, the marginalized
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{multline}\label{LEeq1}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \\ \times \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
where $x_N$ denotes the loudness of the loudest event, and $R_f$ and
$R_b$ are the number of events expected \emph{above our original
  threshold} (so, for example, $R_f (1 - \hat{F}(x_N,\theta))$ is the
number of foreground events expected above loudness $x_N$).  In the
loudest event statistic paper~\citep{Biswas2009}, the authors assume
the background distribution and rate are known, which corresponds to
using a delta function prior on $R_b$. They further assume a flat
prior on $R_f$ and that the foreground and background distributions do
not depend on any unknown free parameters. With these assumptions, the
posterior on $R_f$, Eq.~\eqref{LEeq1}, is modified to
\begin{multline}
p_{\rm LE}(R_f|d) \propto  \left(R_f
\hat{f}(x_N) + R_b \hat{b}(x_N)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N)) + R_b (1-\hat{B}(x_N)) \right)
  \right] .
\end{multline}
Integrating over $R_f$ gives
\begin{multline}
\int_0^\infty p_{\rm LE}(R_f|d) {\rm d}R_f =  \frac{R_b \hat{b}(x_N)}{(1-\hat{F}(x_N))} {\rm e}^{-(1-\hat{B}(x_N)) R_b}\\
%\exp\left[-(1-\hat{B}(x_N))R_b\right] \\ 
\times \left(\frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)} + 1 \right)
\end{multline}
and so the normalised posterior is
\begin{multline}
p_{\rm LE}(R_f|d) = \frac{(1-\hat{F}(x_N))}{1+\Lambda} \left( 1+ R_f (1-\hat{F}(x_N))\Lambda \right) \\
\times \exp\left[-R_f (1-\hat{F}(x_N))\right]
\label{loudest-eq}
\end{multline}
in which we have defined
\begin{equation}
\Lambda \equiv \frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)}.
\end{equation}
With the further identification $\mu \equiv R_f$ and $\hat{\epsilon} \equiv 1 - \hat{F}(x_N)$, this is Eq.~(18) of~\citep{Biswas2009} and we have shown how their parameter $\Lambda$ is related to the foreground and background distributions used here.

@@

If $\tilde{b}(x_N,\theta) \ll \tilde{f}(x_N,\theta)$, we obtain the result $(1-
\hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be understood as the
statement that the rate of foreground events with ranking statistic
greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$, is of order $1$,
as expected. However, $\tilde{b}(x_N,\theta) = -{\rm d}
[\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term may be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) \gg \tilde{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavored.

@@

\section{Thresholding}
This paper is concerned with Bayesian rate estimates based on lists of
events.  Ideally, the lists should contain all events in the data set.
However, for experimental or computational reasons one may wish to
restrict the events to only those above some loudness threshold; in
some cases the rate of foreground or background events, or both, is
even expected to diverge at certain loudnesses.  In this subsection we
address the question of how the rate estimate depends on the threshold
value.

@@

Suppose we attempt to detect gravitational wave signals in a data
stream by matched filtering in the frequency domain against a set of
$N$ template waveforms \citep[e.g.,][]{findchirppaper,LVC2011}.  In
our simplistic model, we suppose the data stream consists of
stationary Gaussian noise with a power spectral density $S(f)$
combined additively with some number of gravitational wave signals.
We assume that the signals are sufficiently rare that they do not
overlap in the data stream.  The signal-to-noise ratio (SNR) of a
template, $h(f)$, given data, $d(f)$, is
\begin{equation}
  \rho_h \equiv \frac{\left\langle h, d \right\rangle}{\sqrt{\left
      \langle h, h \right\rangle}},
\end{equation}
where $\left \langle \cdot \right\rangle$ denotes the noise-weighted
inner product:
\begin{equation}
  \left\langle a, b \right\rangle \equiv 4 \Re \int_0^\infty df\,
  \frac{a^*(f) b(f)}{S(f)}.
\end{equation}
We suppose for simplicity that the templates are sufficiently distinct
that
\begin{equation}
  \left\langle h_i, h_j \right\rangle \simeq \delta_{ij}.
\end{equation}
In the following subsection, we will generalize the model to
overlapping templates.  We rank candidate events by their maximum SNR
over the entire template bank,
\begin{equation}
  x \equiv \max_{h} \rho_h,
\end{equation}
and consider only events that have a maximum SNR above some threshold,
$x > \xmin$.

@@

Through most of this paper, we have assumed that the shape of the
foreground and background distributions is known, or at least can be
modeled with several additional parameters.  This is not necessarily
easy to do.  For example, in the case of gravitational-wave data
analysis, the shape of the foreground distribution of events may
depend on the details of a complex data-analysis pipeline as well as
the astrophysical source distribution, while the background event
distribution depends on data quality and may deviate significantly
from the simple Gaussian-noise behavior modeled in section
\ref{sec:GW-example}.  Several approaches have been developed to
accurately model both distributions, e.g., through the use of injected
signals to model the foreground distribution \cite{ihope,CannonHannaKeppel:2012}.
However, this is a difficult problem (e.g., because of the need to
estimate the background at the very tails of the distribution), and
will require significant future work.  In Section \ref{sec:uncertain-dist}, 
we discussed some of the possible
approaches when the shapes of the background and foreground distributions
cannot be confidently described by models with a few adjustable parameters.

A further complication is that we have considered the rate of events
in the data as products of some analysis pipeline.  This rate may be
different from the physical rate of interest, such as the rate of
compact-binary mergers per unit time per unit volume which generate
gravitational waves, or the physical numbers of stars in the cluster
and field populations which produce the observed luminosities.  Again,
the conversion between the two will depend on the details of the
data-analysis algorithm and ranking statistic, including any selection
effects, and would need to be determined on a case-by-case basis.

>>>>

The task of estimating rates of events when a mixture of foreground
and background events is present in data is a common one in physical
and astrophysical applications.  This problem comes up, among others,
in gravitational-wave data analysis
\cite[e.g.,][]{s5-lowmass,s5-highmass,s5-highmass-erratum,s5-lowmass-2,s6-lowmass,s6-highmass}
and in astronomical observations of a field of objects of mixed
provenance \cite{DeGennaro2009}.  In this paper, we introduce a robust
formalism for estimating event rates from the data when the shape of
foreground and background distributions are known (or parameterized),
but the provenance of individual events as either background or
foreground is unknown.

@@

Two alternative approaches to rate estimation have been suggested and
are commonly used.  One, known as the \emph{loudest-event statistic}
\cite{Biswas2009,Brady2004,Brady2008}, uses only the information from
the highest-ranked event in the data to infer the rate distribution.
This approach has been used successfully
\cite{s5-lowmass,s5-lowmass-2,s5-highmass,s5-highmass-erratum,s6-lowmass,s6-highmass}
when the number of loud foreground events is small (typically zero or
one) to obtain upper limits on foreground rates.  However, the
loudest-event statistic ignores all events except the loudest one, and
so suffers from an unnecessary loss of information; therefore, we
expect it to yield a much larger variance than strictly necessary when
multiple events are present in the data.  In practice, the
loudest-event statistic is typically applied repeatedly to multiple
``chunks'' of data, using the estimated rate posterior from each chunk
as a rate prior for the next chunk's analysis
\cite{s5-lowmass-2,s5-highmass,s5-highmass-erratum}.  Even when used
in this mode, the method discards information, with the amout of
information loss depending on the (arbitrary) division of the data
into chunks.

@@

Ref.~\cite{Messenger2012} considered the problem of determining an
intrinsic rate and population parameters in the presence of missing
data, either due to thresholding, poor sensitivity, or contamination
from noise events.  The approach is complementary to ours: we consider
the problem of accurately counting the events of different classes
present in a dataset, while Ref.~\cite{Messenger2012} deals with
translating such counts into physical rates by properly accounting for
the selection effects on the data set.

@@

For the usual $k=1$ case \citep{Biswas2009}, the marginalized
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{multline}\label{LEeq1}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \\ \times \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
where $x_N$ denotes the loudness of the loudest event, and $R_f$ and
$R_b$ are the number of events expected \emph{above our original
  threshold} (so, for example, $R_f (1 - \hat{F}(x_N,\theta))$ is the
number of foreground events expected above loudness $x_N$).  In the
loudest event statistic paper~\citep{Biswas2009}, the authors assume
the background distribution and rate are known, which corresponds to
using a narrow prior on $R_b$. They further assume a flat prior (in
the absence of other experimental data) on $R_f$ and that the
foreground and background distributions do not depend on any unknown
free parameters. With these assumptions, the posterior on $R_f$,
Eq.~\eqref{LEeq1}, is modified to
\begin{multline}
p_{\rm LE}(R_f|d) \propto  \left(R_f
\hat{f}(x_N) + R_b \hat{b}(x_N)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N)) + R_b (1-\hat{B}(x_N)) \right)
  \right] .
\end{multline}
Integrating over $R_f$ gives
\begin{multline}
\int_0^\infty p_{\rm LE}(R_f|d) {\rm d}R_f =  \frac{R_b \hat{b}(x_N)}{(1-\hat{F}(x_N))} {\rm e}^{-(1-\hat{B}(x_N)) R_b}\\
%\exp\left[-(1-\hat{B}(x_N))R_b\right] \\ 
\times \left(\frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)} + 1 \right)
\end{multline}
and so the normalised posterior is
\begin{multline}
p_{\rm LE}(R_f|d) = \frac{(1-\hat{F}(x_N))}{1+\Lambda} \left( 1+ R_f (1-\hat{F}(x_N))\Lambda \right) \\
\times \exp\left[-R_f (1-\hat{F}(x_N))\right]
\label{loudest-eq}
\end{multline}
in which we have defined
\begin{equation}
\Lambda \equiv \frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)}.
\end{equation}
With the further identification $\mu \equiv R_f$ and $\hat{\epsilon} \equiv 1 - \hat{F}(x_N)$, this is Eq.~(14) of~\citep{Biswas2009} and we have shown how their parameter $\Lambda$ is related to the foreground and background distributions used here.

@@

If $\tilde{b}(x_N,\theta) \ll \tilde{f}(x_N,\theta)$, we obtain the
result $(1- \hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be
understood as the statement that the rate of foreground events with
ranking statistic greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$,
is of order $1$, as expected. However, $\tilde{b}(x_N,\theta) = -{\rm
  d} [\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term may be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) \gg \tilde{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavored.  As we will see, the problem can also be avoided
in the context of the loudest-event framework by even very weak prior
information on the background rate, $R_b$, of the kind present in
nearly all experiments.

@@

\section{Thresholding}
This paper is concerned with Bayesian rate estimates based on lists of
events.  Ideally, the lists should contain all events in the data set.
However, for experimental or computational reasons one may wish to
restrict the events to only those above some loudness threshold; in
some cases the rate of foreground or background events, or both, is
even expected to diverge at certain loudnesses.  In this subsection we
address the question of how the rate estimate depends on the threshold
value.  For a discussion of selection effects, of which thresholding
is but one, on the estimate of physical rates, see
Ref.~\cite{Messenger2012}.

@@

Suppose we attempt to detect gravitational wave signals in a data
stream by matched filtering in the frequency domain against a set of
$N$ template waveforms \citep[e.g.,][]{findchirppaper,s6-lowmass}.  We
use an extremely simplified model of such a search and the ensuing
analysis to demonstrate how our framework could be used in practice.

In our simplistic model, we suppose the data stream consists of
stationary Gaussian noise with a power spectral density $S(f)$
combined additively with some number of gravitational wave signals.
We assume that the signals are sufficiently rare that they do not
overlap in the data stream.  The signal-to-noise ratio (SNR) of a
template, $h(f)$, given data, $d(f)$, is
\begin{equation}
  \rho_h \equiv \frac{\left\langle h, d \right\rangle}{\sqrt{\left
      \langle h, h \right\rangle}},
\end{equation}
where $\left \langle \cdot \right\rangle$ denotes the noise-weighted
inner product:
\begin{equation}
  \left\langle a, b \right\rangle \equiv 4 \Re \int_0^\infty df\,
  \frac{a^*(f) b(f)}{S(f)}.
\end{equation}
We suppose for simplicity that the templates are sufficiently distinct
that
\begin{equation}
  \left\langle h_i, h_j \right\rangle \simeq \delta_{ij}.
\end{equation}
In the following subsection, we will generalize the model to
overlapping templates.  We rank candidate events by their maximum SNR
over the entire template bank,
\begin{equation}
  x \equiv \max_{h} \rho_h,
\end{equation}
and consider only events that have a maximum SNR above some threshold,
$x > \xmin$.

@@

Through most of this paper, we have assumed that the shape of the
foreground and background distributions is known, or at least can be
modeled with several additional parameters.  This is not necessarily
easy to do.  For example, in the case of gravitational-wave data
analysis, the shape of the foreground distribution of events may
depend on the details of a complex data-analysis pipeline as well as
the astrophysical source distribution, while the background event
distribution depends on data quality and may deviate significantly
from the simple Gaussian-noise behavior modeled in section
\ref{sec:GW-example}.  Several approaches have been developed to
accurately model both distributions, e.g., through the use of injected
signals \cite{ihope} or other methods \cite{CannonHannaKeppel:2012} to
model the foreground distribution.  However, this is a difficult
problem (e.g., because of the need to estimate the background at the
very tails of the distribution), and will require significant future
work.  In Section \ref{sec:uncertain-dist}, we discussed some of the
possible approaches when the shapes of the background and foreground
distributions cannot be confidently described by models with a few
adjustable parameters.

A further complication is that we have considered the rate of events
in the data as products of some analysis pipeline.  This rate may be
different from the physical rate of interest, such as the rate of
compact-binary mergers per unit time per unit volume which generate
gravitational waves, or the physical numbers of stars in the cluster
and field populations which produce the observed luminosities.  Again,
the conversion between the two will depend on the details of the
data-analysis algorithm and ranking statistic, including any selection
effects \cite{Messenger2012}, and would need to be determined on a
case-by-case basis.  See Ref.~\cite{Brady2008} for an example of such
conversion when the underlying framework is the loudest-event
statistic.
